Issues in developing multivariable molecular signatures for guiding clinical care decisions

### 2016-07-05 Documents/paper-revised-round2.tex
revisions

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{June 30, 2016}

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

@@

@@

Omics technologies that generate a large amount of molecular data
characterizing patient biospecimens have the potential to provide
information about patients' disease characteristics above and beyond
standard clinical and pathological features. By combining the
information from a large number of molecular features into a
multivariable model or decision algorithm, called a biomarker signature,
there is the opportunity to identify distinct subgroups of patients for
whom treatment decisions can be personalized. A biomarker signature can
guide decisions to treat or not to treat and help identify the patients
who are most likely to have a more favororable outcome or benefit from a
particular therapy. The key challenge is to combine features from a high
dimensional molecular assay to derive a signature with good clinical
performance and appropriately characterize its performance. The
inappropriate practice of using overlapping data to both build a
signature and evaluate its performance can lead to severe over-optimism
bias in performance estimates. We summarize the key statistical issues
and methods for developing and validating biomarker signatures, using
examples from the literature for illustration.

@@

Let \(\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}\) denote
the statistic that quantifies the performance of the function \(f\),
such as predictive accuracy, mean squared error, or area under the
receiver operating characteristic (ROC) curve (AUC). This could also be
a measure of association, such as an odds ratio, hazard ratio, or
log-rank statistic. This is a function of both \(f\) and \(S\). We are
interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
expected error under the data generation mechanism (distribution
\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\). This allows
us to understand how the signature will perform on future observations
generated from \(\mathcal{P}\). We may also be interested in estimating
\(E_\mathcal{P}[\phi_f(S)]\) \textbf{for all} \(f \in \mathcal{F}\),
which is the generalization performance for \(f\) generated using
mechanism \(\mathcal{F}\). This doesn't guide outside researchers as to
which specific \(f\) to use, yet it is useful for development because it
tells us how much signal is in the data. As shorthand we will write this
as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).

@@

Our focus here will be mainly on the development of signatures using
supervised methods, meaning methods that use information on an outcome
variable it is desired to predict during the signature development
process. Regression methods are a commonly used class of supervised
methods. Oncotype DX \citep{paik2004multigene} is an example of a
prognostic signature used for clinical decision making in early stage
breast cancer that was developed using supervised methods. In the case
of Oncotype DX, the outcome \(Y\) was time to distant recurrence of
breast cancer and the feature data were gene expression values measured
in breast tumor specimens. A common supervised approach to identifying
treatment-selection signatures is to use regression techniques to
estimate a signature that has a strong interaction with a particular
treatment. Supervised methods are contrasted with unsupervised methods
which use only the feature data to derive a signature. An example of a
signature developed by unsupervised methods is the signature that
identifies biological subclasses of diffuse large B-cell lymphoma, which
were originally developed using clustering methods and subsequently were
found to be associated with prognosis \citep{alizadeh2000distinct}. It
is possible, and quite common in high-dimensional data settings, to
combine multiple approaches when estimated a signature. For instance, a
data-reduction step by variable selection or clustering may be performed
before doing regression analysis on the resulting components. A review
by Subramanian and Simon identified a large collection of gene
expression--based prognostic signatures in lung cancer which had been
developed using a wide variety of methods \citep{subramanian2010gene}.

The focus of this paper is how to obtain a valid estimate of signature
performance as quantified by a metric \(\phi\), that is, a good estimate
of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
apply no matter what signature development method is used. Performance
depends on the true signal in the data and the specific algorithm
\(\mathcal{F}\) used for development. The performance will reflect that
expected of signatures developed applying the prescribed methods to data
drawn from the distribution \(\mathcal{P}\). For signature deployment,
one would also want a particular specification of \(f\) for others to
use on independent data (e.g., in clinical practice). It is not trivial
to conduct this evaluation in a valid manner when there is a limited
data set available to both define a signature and assess its
performance. We illustrate the potential for bias to creep into the
performance estimation and review some strategies to avoid the pitfalls
that lead to these biases. We evaluate various strategies through a
simulation study and compare the performance of some of those strategies
using a real data exmaple involving the development of a signature for
use in making treatment decisions for patients with lung cancer.

@@

The main goal of interest here is to estimate
\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
performance metric on future observations for \(f \in \mathcal{F}\).
This can be estimated with the in-sample empirical estimate:
\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
particular \(f\). However, if \(S\) is used to estimate \(f\) then the
estimate will be biased due to overfitting, that is,
\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large.
Overfitting occurs when a model is fit to noise in the data. This often
occurs when fitting a model that is overly complex relative to the
amount of signal in the available data. Overfitting bias results from
the fact that \(\phi_f(S)\) depends on \(f\) which depends on \(S\), and
thus the statistic \(\phi_f(S)\) is being adaptively defined based on
the observed data \(S\). Any estimate of signature performance is
potentially subject to bias if overfitting occurs, but such biases
frequently go unrecognized in the medical literature.

Typically a performance metric \(\phi\) used during signature
development is degined to quantify either calibration or discrimination,
or some combination of the two. Calibration assesses how well the
signature based predictions compare to the observed outcomes.
Discrimination assesses how well the signature distinguishes between
groups of patients who do or do not experience an event. To assess
discrimination, a different metric \(\phi\) may be used, such as the
area under the ROC curve. Measures of association such as the odds
ratio, hazard ratio, or difference in survival probabilities are also
commonly used performance metrics, although their value for biomarker
signatures has been debated \citep{pepe2004limitations}. In
\citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing the
high risk and low risk groups. The risk groups were determined by the
signature \(f\), which was estimated using the JBL.10 observation arm
data. The signature was then applied to the same data that were used to
build it to define risk groups, producing a the hazard ratio estimate of
15 for a prognostic effect in the observation arm. Subsequent
evaluations of the signature's performance on independent data resulted
in hazard ratio estimates in the 2 - 3 range, substantially smaller than
the original estimate of 15. Next we reanalyze the JBR.10 data and
illustrate some methods to avoid bias in evaluation of signature
performance.

@@

A traditional method to avoid overfitting is the split sample approach.
First, randomly partition \(S\) into the training sample \(S_t\) and the
holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
respectively. Then, \(S_h\) is hidden from the analyst while
\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
estimator of \(E_{\mathcal{P}}[\phi_{f_t}(S)]\).
\citet{dobbin2011optimally} investigate how to optimally split a dataset
into training and holdout partitions. The specific form of \(f_t\) that
is fixed using \(S_t\) can be reported as the function for others to
use, therefore the aforementioned estimator applies for that specific
\(f_t\). The drawback of the split-sample approach is that the
performance of \(f_t\) is likely to be inferior to the performance of a
function \(f^*\) derived using the same approach applied to the entire
data set. \citet{dobbin2007sample} investigate how to choose the sample
size n so that there is high probability that the signature developed
will have expected performance within some specified margin of the the
optimal signature that could be developed if the development data set
had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a
resampling based approach. For a fixed integer \(k\), which can be
between 1 and \(n\), we randomly partition the full data set \(S\) into
subgroups of size \(k\) (assume for simplicity here that \(n\) is evenly
divisible by \(k\) so that \(n/k\) subgroups are formed). For each
\(k\), \(f_{-k}\) is estimated and fixed by appling \(\mathcal{F}\) on
\(S_{-k}\) which is the subset of \(S\) that is disjoint from \(S_k\).
Then, we get an estimate \(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an
estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is
repeated \(K\) times to yield \(K\) estimates and then these are
averaged to obtain the performance estimate. This process is called
``leave \(k\) out'' or ``\(n/k\) fold'' cross-validation. Note that for
each subgroup in the partition, we obtain a new estimate of \(f\),
therefore we are only estimating
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
single specified signature. Typically, if a specific form for \(f\) is
desired, it would be estimated using the entire dataset \(S\) to yield
\(f^*\) as above. Importantly one should report the cross-validated
estimate just described and not the naive esitimate that would be
obtained by substituting the full data \(S\) into the performance metric
calculation for the signature \(f^*\). The problematic aspects of such
naive ``resubstitution estimates'' of performance are discuss in more
detail below.

A variation on the cross-validation approach is bootstrapping. In that
case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
from \(S\). Then \(f_{b}\) is estimated and fixed by applying
\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
calculated on the subset of \(S\) that is disjoint from \(S_b\), which
we denote by \(S_{-b}\), to yield an estimate of
\(E[\phi_{f_b}(S_{-b})]\). This process is repeated \(K\) times to yield
\(K\) estimates. These \(K\) performance metric estimates are averaged
to obtain the mean over the bootstrap replicates.
\citet{efron1997improvements} suggest a variation, the 0.632 estimate:
\[
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
using the entire dataset.

Another variation on all of these methods is the concept of
pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
of computing the statistic \(\phi\) for each of the held-out subsets
(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
fitted signature \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
to obtain a set of pre-validated signature estimates \(\hat{f}\) which
are then used to calculate \(\phi\). For single-step split sample
training and validation, this process is equivalent to what is described
above. For cross-validation and the bootstrap, this process avoids the
problem of having too few cases to estimate the statistic \(\phi\) on
each of the smaller held-out datasets.

@@

We compare each of the methods described above: split-sample holdout,
cross-validation, bootstrap, and pre-validation, along with several
commonly used but biased approaches. The biased approaches are described
in Table \ref{descript}. Two of the biased approaches use the full
sample to select the features, followed by fitting the multivariable
model on the holdout subset. This is referred to as ``parital holdout''
or ``partial CV'' when using split-sample holdout or cross-validation as
the validation step, respectively. We also implemented the naïve
resubstitution approach, wherein the model is trained and evaluated on
the same dataset, and the partial resubstitution approach wherein the
model is developed on a training subset but then evaluated on the
combined training adn holdout data sets. Our main interest is in
comparing the bias and variance of the resulting estimates of
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
at two different performance metrics, the area under the ROC curve (AUC)
and the odds ratio for the outcome comparing the signature groups.

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are \(N\) observations
and \(p\) features. The true value of the AUC is 0.5. CV = Cross
validation. \label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are \(N\) observations and \(p\)
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}

@@

\citet{zhu2010prognostic} present results that mainly focus on the
association of their signature with outcome, albeit a less than ideal
measure of discrimination ability. They demonstrate that the two risk
subgroups predicted by their signature (high risk and low risk) have
separation in their survival curves and that the hazard ratio for their
signature is large and significant even when adjusting for other risk
factors. They do not directly address calibration, that is, whether
their signature accurately predicts survival times.

@@

The signature development procedure was described in the introduction.
There is both a feature selection step, and a multivariable estimation
step. This results in a continuous signature which is the linear
predictor of a Cox regression model. The signature is dichotomized by
selecting the cutoff that yields the most significant log-rank statistic
for comparing the resulting risk groups. Discrimination of the signature
is assessed using the concordance statistic as implemented in the
\texttt{survival} package in \texttt{R} \citep{survival}. To paraphrase
the help file: this is defined as the probability of agreement for any
two randomly chosen observations, which in this case means that the
observation with the shorter survival time also has the larger signature
value. This is similar to an interpretation of the AUC for binary data.

First we fit the signature using the entire observation cohort (n = 62).
The signature was then evaluated on the same dataset. The survival plot
on the left side of Figure \ref{fig2} shows extreme separation between
the two risk groups (HR = 20, p \textless{} 0.001), consistent with the
reported JBL.10 signature, and the estimated concordance is 0.87. After
correctly accounting for the selection process, our estimates of
association and discrimination are much less impressive.

The right plot in Figure \ref{fig2} shows the survival curves for the
two risk groups using the pre-validated estimates of the risk score. We
partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
Then for each group \(b\), we fit the model using \(S_{-b}\) and
obtained prevalidated estimates for \(S_{b}\). The survival curves plot
the survival times for comparing risk groups using the prevalidated
estimates. The separation is much less impressive. The concordance
between the prevalidated signature and the survival times is 0.61,
indicating much worse discrimination.

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/examppresub-1.pdf}
\caption{Comparison of survival by gene expression based risk signature.
The left plot shows the resubstitution estimate, while the right plot
shows the pre-validated estimate. \label{fig2}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/inter1-1.pdf}
\caption{Survival curves comparing the treatment effect by gene
expression signature risk group. The top set of plots shows the partial
resubstitution based signature and the bottom row shows the
pre-validated signature estimates. \label{fig3}}
\end{figure}

\begin{longtable}[c]{@{}lllll@{}}
\caption{Hazard ratios and 95\% confidence intervals from separate Cox
regression models that adjust for tumor histologic subtype, stages, age,
and sex. Rows labeled `High risk vs low risk' show the hazard ratio for
the signature-based risk group comparison. The rows labeled `Trt/Risk
interaction' show the hazard ratio for the interaction term of treatment
by signature-based risk group. The partial substitution estimates are
dramatically optimistically biased. \label{adjhr}}\tabularnewline
\toprule
Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
\midrule
\endfirsthead
\toprule
Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
\midrule
\endhead
\textbf{Partial Resubstitution} & High Risk vs Low Risk & 38.9 & 9.2 to
164.7 & \textless{} 0.001\tabularnewline
& Trt/Risk interaction & 14.7 & 3.2 to 67.0 & \textless{}
0.001\tabularnewline
\textbf{Prevalidation} & High Risk vs Low Risk & 1.9 & 0.8 to 4.3 &
0.122\tabularnewline
& Trt/Risk interaction & 1.8 & 0.5 to 6.5 & 0.395\tabularnewline
\bottomrule
\end{longtable}

>>>>

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs$^*$ and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{July 05, 2016}

\usepackage{setspace}
\usepackage{tablefootnote}
\doublespacing

@@

\begin{center}
\textit{Biostatistics Branch, Biometric Research Program \\
National Cancer Institute \\
9609 Medical Center Drive, MSC 9735 \\
Bethesda, MD 20892-9735 \\
Phone: 240-276-6004 \\
Email: michael.sachs@nih.gov \\
*: Corresponding Author}

\end{center}

@@

Omics technologies that generate a large amount of molecular data
characterizing biospecimens have the potential to provide information
about patients' disease characteristics beyond standard clinical
features. By combining information from a large number of features into
a multivariable model, called a biomarker signature, there is the
opportunity to identify distinct subgroups of patients for whom
treatment decisions can be personalized. The key challenge is to derive
a signature with good performance and appropriately characterize its
performance. We summarize the key statistical issues and methods for
developing and validating biomarker signatures, using examples from the
literature for illustration. \\

\textbf{keywords}: biomarkers; biomarker signatures; personalized medicine

\clearpage

@@

Let \(\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}\) denote
the statistic that quantifies the performance of the function \(f\).
This is a function of both \(f\) and \(S\). We are interested in
estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the expected
performance under the data generation mechanism (distribution
\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\) when \(f^*\)
is applied to \(X\) for the purpose of predicting \(Y\). This
expectation is with respect to \(\mathcal{P}\), the distribution of the
random variable \(S\). This allows us to understand how the signature
will perform on future observations generated from \(\mathcal{P}\). The
difficulty in estimating this quantity is that generally only a single
sample from \(\mathcal{P}\) is available to both develop \(f^*\) and to
estimate its expected performance \(E_\mathcal{P}[\phi_{f^*}(S)]\).

We may also be interested in estimating \(E_\mathcal{P}[\phi_f(S)]\)
\textbf{for all} \(f \in \mathcal{F}\), which is the generalization
performance for \(f\) generated using mechanism \(\mathcal{F}\). This
doesn't guide outside researchers as to which specific \(f\) to use, yet
it is useful for development because it tells us how much signal is in
the data and how well it can be captured with \(f \in \mathcal{F}\). As
shorthand we will write this as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).

@@

Our focus here will be mainly on the development of signatures using
supervised methods, meaning methods that use information about an
outcome variable that it is desired to predict. Regression methods are a
commonly used class of supervised methods. Oncotype DX
\citep{paik2004multigene} is an example of a prognostic signature used
for clinical decision making in early stage breast cancer that was
developed using supervised methods. In the case of Oncotype DX, the
outcome \(Y\) was time to distant recurrence of breast cancer and the
feature data were gene expression values measured in breast tumor
specimens. Regression methods can also be used to develop
treatment-selection (also called predictive) signatures that exhibit a
strong interaction with a particular treatment.

Supervised methods are contrasted with unsupervised methods which use
only the feature data to derive a signature. An example of a signature
developed by unsupervised methods is the signature that identifies
biological subclasses of diffuse large B-cell lymphoma, which were
originally developed using clustering methods and subsequently were
found to be associated with prognosis \citep{alizadeh2000distinct}. It
is possible, and quite common in high-dimensional data settings, to
combine multiple approaches when estimating a signature. For instance, a
data-reduction step by variable selection or clustering may be performed
before doing regression analysis on the resulting components. A review
by Subramanian and Simon identified a large collection of gene
expression--based prognostic signatures in lung cancer which had been
developed using a wide variety of methods \citep{subramanian2010gene}.

The focus of this paper is how to obtain a valid estimate of signature
performance as quantified by a metric \(\phi\), that is, a good estimate
of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
apply no matter what signature development method is used. Performance
depends on the true signal in the data and the specific algorithm
\(\mathcal{F}\) used for development. The performance will reflect that
expected of signatures developed applying the prescribed methods to data
drawn from the distribution \(\mathcal{P}\). For signature deployment,
one would also want a particular specification of \(f\) for others to
use on independent data (e.g., in clinical practice). It is not trivial
to conduct this evaluation in a valid manner when there is a limited
data set available to both define a signature and assess its
performance. We illustrate the potential for bias to creep into the
performance estimation and review some strategies to avoid the pitfalls
that lead to those biases. We evaluate various strategies through a
simulation study and compare the performance of some of those strategies
using a real data exmaple involving the development of a signature for
use in making treatment decisions for patients with lung cancer. All
simulations and analyses were conducted using \texttt{R} version 3.2.4
\citep{arr}.

@@

\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\) can be estimated with the
in-sample empirical estimate:
\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
particular \(f\). However, if \(S\) is used to estimate \(f\) then the
estimate will be biased due to overfitting, that is,
\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large.
Overfitting occurs when a model is fit to noise in the data. This often
occurs when fitting a model that is overly complex relative to the
amount of signal in the available data. Overfitting bias results from
the fact that \(\phi_f(S)\) depends on \(f\) which depends on \(S\), and
thus the statistic \(\phi_f(S)\) is being adaptively defined based on
the observed data \(S\). Any estimate of signature performance is
potentially subject to bias if overfitting occurs, but such biases
frequently go unrecognized in the medical literature.

Typically a performance metric \(\phi\) used during signature
development is designed to quantify either calibration or
discrimination, or some combination of the two. Calibration assesses how
well the signature-based predictions compare to the observed outcomes.
Discrimination assesses how well the signature distinguishes between
groups of patients who do or do not experience an event or have a
particular phenotype. To assess discrimination, a different metric
\(\phi\) may be used, such as the area under the ROC curve. Measures of
association such as the odds ratio, hazard ratio, or difference in
survival probabilities are also commonly used performance metrics,
although their value for biomarker signatures has been debated
\citep{pepe2004limitations}. In \citet{zhu2010prognostic}, \(\phi\) was
the hazard ratio comparing disease-specific survival between the groups
identified as high risk and low risk. The risk groups were determined by
the signature \(f\), which was estimated using the JBL.10 observation
arm data. The signature was then applied to the same data that were used
to build it to define risk groups, producing the hazard ratio estimate
of 15 for the prognostic effect comparing signature-defined risk groups
in the observation arm. Subsequent evaluations of the signature's
performance on independent data resulted in hazard ratio estimates in
the 2 - 3 range, substantially smaller than the original estimate of 15.
This disparity is due, at least in part, to the bias in the original
hazard ratio estimate of 15 resulting from reuse of the data used to
build the signature. Next we discuss some methods to avoid bias in
evaluation of signature performance.

@@

A traditional method to avoid overfitting is the split sample holdout
approach. First, randomly partition \(S\) into the training sample
\(S_t\) and the holdout sample \(S_h\) with sample sizes \(n_t\) and
\(n_h\), respectively. Then, \(S_h\) is hidden from the analyst while
\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
estimator of \(E_{\mathcal{P}}[\phi_{f_t}(S)]\).
\citet{dobbin2011optimally} investigate how to optimally split a dataset
into training and holdout partitions. The specific form of \(f_t\) that
is fixed using \(S_t\) can be reported as the function for others to
use, therefore the aforementioned estimator applies for that specific
\(f_t\). Alternatively, \(f^*\), the signature derived using the entire
set \(S\) can be reported, accompanied by the estimate of the
performance of \(f_t\). The drawback of this approach is that the
performance of \(f_t\) is likely to be inferior to the performance of a
function \(f^*\) derived using the same approach applied to the entire
data set. \citet{dobbin2007sample} investigate how to choose the sample
size n so that there is high probability that the signature developed
will have expected performance within some specified margin of the the
optimal signature that could be developed if the development data set
had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a
resampling based approach. For a fixed integer \(k\), which can be
between 1 and \(n\), we randomly partition the full data set \(S\) into
subgroups of size \(k\) (assume for simplicity here that \(n\) is evenly
divisible by \(k\) so that \(K = n/k\) subgroups are formed). For each
\(k\), \(f_{-k}\) is estimated and fixed by appling \(\mathcal{F}\) on
\(S_{-k}\) which is the subset of \(S\) that is disjoint from \(S_k\).
Then, we get an estimate \(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an
estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). Then one
iterates through the \(K\) subgroups to yield \(K\) estimates and then
these are averaged to obtain the performance estimate. This process is
called ``leave \(k\) out'' or ``\(K\) fold'' cross-validation. Note that
for each subgroup in the partition, we obtain a new estimate of \(f\),
therefore we are only estimating
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
single specified signature. Typically, if a specific form for \(f\) is
desired, it would be estimated using the entire dataset \(S\) to yield
\(f^*\) as above. Importantly one should report the cross-validated
estimate just described and not the naïve estimate that would be
obtained by substituting the full data \(S\) into the performance metric
calculation for the signature \(f^*\). The problematic aspects of such
naive ``resubstitution estimates'' of performance are discussed in more
detail below.

A variation on the cross-validation approach is bootstrapping. In that
case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
from \(S\). Then \(f_{b}\) is estimated and fixed by applying
\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
calculated on the subset of \(S\) that is disjoint from \(S_b\), which
we denote by \(S_{-b}\), to yield an estimate of
\(E[\phi_{f_b}(S_{-b})]\). This process is repeated \(K\) times to yield
\(K\) estimates. These \(K\) performance metric estimates are averaged
to obtain the mean over the bootstrap replicates.
\citet{efron1997improvements} suggest a variation, the 0.632 estimate:
\[
\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
\] where \(\hat{E}[\phi_{f}(S)]\) is the naïve resubstitution estimate
of \(\phi_f\) using the entire dataset.

Another variation on all of these methods is the concept of
pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
of computing the statistic \(\phi\) for each of the held-out subsets
(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
fitted signature \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
to obtain a set of pre-validated signature estimates
\(\{\hat{f}_i, i = 1, \ldots, n\}\). Then some measure of the
association between these signature estimates and the withheld true
outcomes, \(Y_i, i = 1, \ldots, n\), is estimated to serve as a measure
of the signature performance. The association measure \(\phi\) could be
estimated as a coefficient in a linear, Cox proportional hazards, or
logistic regression model, with or without adjustment for other
covariates, as appropriate for the situation. The idea behind this
pre-validation approach is that if the signature has good performance,
then it should produce values that are strongly associated with the true
outcomes. In contrast to cross-validation and the bootstrap, this
process avoids the problem of having too few cases to estimate the
statistic \(\phi\) on each of the smaller held-out datasets.

These resampling-based procedures described above provide unbiased
estimates of predictive accuracy for the procedure, that is
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\), as opposed to the accuracy
for a particular signature that others can use. Often, we will provide
the signature developed on the entire dataset as the one for future use,
as using the full data will estimate signature coefficients with the
most precision. Still, it is important to avoid the temptation to report
the estimate of performance on this full data signature, as it can be
severly biased.

@@

We compare each of the methods described above: split-sample holdout,
cross-validation, bootstrap, and pre-validation, along with several
commonly used but biased approaches. The biased approaches are described
in Table \ref{descript}. Two of the biased approaches use the full
sample to select the features, followed by fitting the multivariable
model on holdout subsets. This is referred to as ``parital holdout'' or
``partial CV'' when using split-sample holdout or cross-validation as
the validation step, respectively. We also implemented the naïve
resubstitution approach, wherein the model is trained and evaluated on
the same dataset, and the partial resubstitution approach wherein the
model is developed on a training subset but then evaluated on the
combined training and holdout data sets. Our main interest is in
comparing the bias and variance of the resulting estimates of
\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
at two different performance metrics, the area under the ROC curve (AUC)
and the odds ratio for the binary outcome comparing the signature
groups.

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-round2_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are \(N\) observations
and \(p\) features. The true value of the AUC is 0.5. CV = Cross
validation. \label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-round2_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are \(N\) observations and \(p\)
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}

@@

\citet{zhu2010prognostic} present results that mainly focus on the
association of their signature with outcome as quantified by a hazard
ratio, albeit a less than ideal measure of discrimination ability. They
demonstrate that the two risk subgroups predicted by their signature
(high risk and low risk) have separation in their survival curves and
that the hazard ratio for their signature is large and significant even
when adjusting for other risk factors. They do not directly address
calibration, that is, whether their signature accurately predicts
survival times.

@@

Now recall the general framework for signature development that was
described in the introduction. There is both a feature selection step,
and a multivariable estimation step. This results in a continuous
signature which is the linear predictor of a Cox regression model. The
signature is dichotomized by selecting the cutoff that yields the most
significant log-rank statistic for comparing the resulting risk groups.
Discrimination of the signature is assessed using the concordance
statistic as implemented in the \texttt{survival} package in \texttt{R}
\citep{survival}. To paraphrase the help file: the concordance statistic
is defined as the probability of agreement for any two randomly chosen
observations, which in this case means that the observation with the
shorter survival time also has the larger signature value. This is
similar to an interpretation of the AUC for binary data.

First we fit the signature using the entire observation arm (n = 62).
The signature was then evaluated on the same dataset. The survival plot
on the left side of Figure \ref{fig3} shows extreme separation between
the two risk groups (HR = 20, p \textless{} 0.001), consistent with the
reported JBL.10 signature, and the estimated concordance is 0.87. After
correctly accounting for the selection process, our estimates of
association and discrimination are much less impressive.

The right plot in Figure \ref{fig3} shows the survival curves for the
two risk groups using the pre-validated estimates of the risk score. We
partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
Then for each group \(b\), we fit the model using \(S_{-b}\) and
obtained prevalidated estimates for \(S_{b}\). The survival curves plot
the survival times for comparing risk groups using the prevalidated
estimates. The separation is much less impressive. The concordance
between the prevalidated signature and the survival times is 0.61,
indicating much worse discrimination.

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-round2_files/figure-latex/examppresub-1.pdf}
\caption{Comparison of survival by gene expression based risk signature.
The left plot shows the resubstitution estimate, while the right plot
shows the pre-validated estimate. \label{fig3}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-round2_files/figure-latex/inter1-1.pdf}
\caption{Survival curves comparing the treatment effect by gene
expression signature risk group. The top set of plots shows the partial
resubstitution based signature and the bottom row shows the
pre-validated signature estimates. \label{fig4}}
\end{figure}

\begin{longtable}[c]{@{}lllll@{}}
\caption{Hazard ratios and 95\% confidence intervals from separate Cox
regression models that adjust for tumor histologic subtype, stages, age,
and sex. Rows labeled `High risk vs low risk' show the hazard ratio for
the signature-based risk group comparison. The rows labeled `Trt/Risk
interaction' show the hazard ratio for the interaction term of treatment
by signature-based risk group. The partial substitution estimates are
dramatically optimistically biased. \label{adjhr}}\tabularnewline
\toprule
Method & Comparison & Hazard Ratio & 95\% CI$^*$ & Adjusted
p$^*$\tabularnewline
\midrule
\endfirsthead
\toprule
Method & Comparison & Hazard Ratio & 95\% CI$^*$ & Adjusted
p$^*$\tabularnewline
\midrule
\endhead
\textbf{Partial Resubstitution} & High Risk vs Low Risk & 38.9 & 9.2 to
164.7 & \textless{} 0.001\tabularnewline
& Trt/Risk interaction & 14.7 & 3.2 to 67.0 & \textless{}
0.001\tabularnewline
\textbf{Prevalidation} & High Risk vs Low Risk & 1.9 & 0.8 to 4.3 &
0.122\tabularnewline
& Trt/Risk interaction & 1.8 & 0.5 to 6.5 & 0.395\tabularnewline
\bottomrule
& \multicolumn{4}{r}{CI's and p-values based on incorrect inference.} \\
\end{longtable}
